---
title: "Reproducible Quantitative Research for Music Education With R"
subtitle: "2026 Biennial NAfME Music Research and Teacher Education Conference"
output: learnr::tutorial
runtime: shiny_prerendered
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
library(learnr)
library(NAFMErSESSION)
library(QuantPsyc)
library(tidyverse)
library(knitr)
library(rstatix)
library(emmeans)
knitr::opts_chunk$set(echo = TRUE)
options(tutorial.storage = list(
  save_object = function(...) { },
  get_object = function(...) NULL,
  get_objects = function(...) list(),
  remove_all_objects = function(...) { }
))
```

## Welcome!

This is a tutorial that was created to facilitate practicing with R in
the context of R Studio during the 2026 Biennial NAfME Music Research
and Teacher Education Conference. You can download both of these
programs to use "offline" here:

- [R](https://www.r-project.org/)
- [R Studio environment](https://posit.co/download/rstudio-desktop/)

### *I like R mainly because...*

1.  It's FREE
2.  It has an amazingly wide library of tools that constantly grows
3.  You can make beautiful, highly customized graphs
4.  The community of users is very helpful and there are abundant
    resources for learning

## Creating an R Project to Organize your Data Analysis Files

### *R Projects*

Using **R Projects** in R Studio will help you get started and help with
file management (e.g., scripts, data) because an R Project automatically
groups all of your files in a single folder/directory.

Setting one up involves the following navigation:

- File -\> "New Project..."
- Then choose to house your R Project in either a "New Directory" (i.e.,
  a new folder on your computer) or an "Existing Directory" (i.e., a
  folder that already exists on your computer)

The lower right panel of R Studio shows the directory address you
created (or chose) for your R Project and all of the files in that
directory. You should save any data files you need for your analyses in
the same directory.

If you don't work from an R Project, you will have to manage the
directory designations for your files manually ... and that can cause
errors and frustration.

### *R Script*

An **R script** file is where you will create and save the code you
write for your projects.

Opening an R script file involves the following navigation:

- File -\> "New File" -\> "R Script"

If you are working in an R Project when you create your R script file,
the new script file will automatically be organized (i.e., saved) with
the rest of your project files.

## Basic Elements of the R Programming Language

### *Assigning Objects and Storing Them in the Global Environment*

R can store objects in its **global environment**, that is similar to a
short-term memory. You can see any objects currently in the global
environment in the upper right panel of R Studio under the tab titled
"Environment."

You store objects by assigning information to a name using a `<-`
symbol. You can then call the named object or perform operations with
it. See the following examples in the code below:

```{r assign_variables, exercise = TRUE, exercise.eval = FALSE}
# The lines showing the assignments of numbers to x and y will not result in any 
# output. However, R will have stored the values under those designated assignments
# in the R Environment.
x <- 4
y <- 4

# Each of the following lines will produce output.
x
y
x + y

# This line will not produce output, it will only store the sum of the information
# in x and y in "z" in the R Environment.
z <- (x + y)

# This line will produce output.
z/2
```

### *Data Types*

Being aware of the **type** of variables present in your workspace is
critical since it will have major implications for the sorts of analyses
you can conduct and if you will be able to execute your code without
errors.

The four common names for levels of measurement (nominal, ordinal,
interval, ratio) are not the terms used in R. R distinguishes between
variable types using the following terminology:

- Numeric: continuous quantitative values (10.5, 55.3, 787.0)
- Integer: discrete quantitative values with no decimal values (1, 55,
  100) 
- Factor: discrete categories with each category called a "level" (an
  analog to a nominal variable) ("cat", "dog", "fish")
- Ordered factor: discrete categories that are ordered (analogous to an
  ordinal variable) ("never", "sometimes", "always")
- Character: letters with no additional information attached ("k", "R is
  exciting", "FALSE") (referred to as "string" in SPSS)
- Logical: TRUE or FALSE Boolean values
- Complex: for the use of imaginary numbers, indicated by "i"

### *Vectors*

You can also assign a collection of values, i.e., a **vector** of
values, to an object using the **concatenate** function `c()`. The
following code shows the storage of variables of three different data
types: *numeric*, *integer*, and *character*. Note that you will have to
separate values with commas and use parentheses when including
characters.

R will do a pretty good job at guessing what type of data the values in
each variable represent when the values are numbers, but it may not do
what you expect.

```{r assign_vectors, exercise = TRUE, exercise.eval = FALSE}
# As with the code above, the lines showing the assignments of vectors to each 
# name will not result in any output. However, R will have stored the vectors 
# under those designated assignments.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")

# Calling the name of each of the vectors will produce output.
numeric_variable
integer_variable
character_variable
```

As with single value assignments to objects, R will also allow you to
perform operations with stored vectors (or really any kind of data
type). See the following as an example:

```{r operate_vectors, exercise = TRUE, exercise.eval = FALSE}
# First create the example variables.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)

# Then perform vector-based operations
numeric_variable/2
integer_variable*2
```

### *Dataframes*

A **dataframe** is R's term for what most would simply call a data set.
You can create dataframes by assembling vectors of the same length into
a single object using the `data.frame()` function. An example of doing
this is provided below.

```{r assign_dataframe, exercise = TRUE, exercise.eval = FALSE}
# First, we store the variables under designated assignments.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")

# Then you can bind them together into a dataset called "my_data" with three 
# columns (i.e., variables) and 10 rows (i.e., entries).
my_data <- data.frame(numeric_variable, integer_variable, character_variable)

# Calling the "my_data" object shows the entire dataset
my_data
```

You can also examine a data frame using a variety of different *viewing*
functions. See below how the `head()` function returns the first six
rows, the `tail()` function returns the last six rows, and the `str()`
function displays a brief description of the data set along with each
variable by type.

```{r view_dataframe, exercise = TRUE, exercise.eval = FALSE}
# First, we store the variables under designated assignments.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")

# Then you can bind them together into a dataset called "my_data" with three 
# columns (i.e., variables) and 10 rows (i.e., entries).
my_data <- data.frame(numeric_variable, integer_variable, character_variable)

head(my_data)
tail(my_data)
str(my_data)
```

To view a single variable from a dataframe you will need to use a `$`
after the name of the dataframe to specify to R that it should search
there to find the variable. For example: `dataframe_name$variable_name`.

```{r view_dataframe_select, exercise = TRUE, exercise.eval = FALSE}
# Again, these first lines are just creating the dataframe for this example.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")
my_data <- data.frame(numeric_variable, integer_variable, character_variable)

# Use the "$" to call a single variable within a dataframe
my_data$numeric_variable
```

You may have noticed earlier that when using the `str()` function the
three variables in the data frame were summarized as "num" (i.e.,
numeric) or "chr" (i.e., character). R assumed that `integer_variable`
was a numeric variable type even though none of the values had decimal
places. It is often necessary to manually tell R what type of data each
variable represents. For example, we can specify that the
`integer_variable` is an integer type variable with the `as.integer()`
function and that the `character_variable` is a factor type variable
with the `as.factor()` function. You can use the `class()` function to
identify the type of data the variable represents (or the "type" of any
object in R, for that matter).

We can replace the original variable in the dataframe with the new one
by using an object assignment approach. Alternatively, we could ask for
R to create a new variable in the dataframe to represent the newly
specified variable while retaining the original. For example:

```{r view_dataframe_specify, exercise = TRUE, exercise.eval = FALSE}
# Again, these first lines are just creating the dataframe for this example.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")
my_data <- data.frame(numeric_variable, integer_variable, character_variable)

# Assign the newly specified variable to the same name as the original to replace it.
my_data$integer_variable <- as.integer(my_data$integer_variable)

# Assign the newly specified variable to a new name to keep the original too.
my_data$character_variable_now_factor <- as.factor(my_data$character_variable)

# View the structure of the data again to see the newly specified variables.
str(my_data)
```

## *Functions and Arguments*

R comes "preloaded" with many, many functions. That is what is referred
to as "base" R.

A function is essentially a self-contained bundle of code stored in a
single name that can be used to execute the whole bundle on a given
object (or objects) at once. For example, the `data.frame()`, `head()`,
`tail()`, and `str()` functions used in the previous section take a
dataframe object and execute some length of code to describe the
contents of that dataframe object. In contrast, we used the
`as.integer()` function above to take a single vector object and execute
some length of code to coerce a numeric object into an integer class
object.

The elements that are put inside the parentheses of a function are
called **arguments**, which are specifications for options regarding how
a given function is used. There are at least three easy way to learn
about the specific arguments possible for each function:

- Calling the name of the function preceded by a `?`, such as
  `?function.name()`
- Looking up the function's description in package documentation on the
  official Comprehensive R Network website
  ([CRAN](https://cran.r-project.org/web/packages/available_packages_by_date.html))
- Searching the "function name" and "R" on the web and looking for the
  many, many ways individual users have explained how to use the
  function (the R community of users is extremely generous and helpful
  when it comes to learning how to use R)

All functions come with default settings for its arguments and although
in most cases those settings will be exactly what you need, you will
eventually need to adjust certain arguments to avoid errors in your
code. For example, you can calculate the arithmetic mean of a variable
using the `mean()` function available in the base setup of R. Assuming
you have no missing data in your variable, calling `mean(variable)` will
work just fine. However, if there are missing values, then the function
will report an error. That's because the `mean()` function has three
arguments and one of them assumes that there are no missing values in
the object you are using it on (see [mean
documentation](https://rdrr.io/r/base/mean.html)).

The arguments for `mean()` are:

- `x` the name of the object to calculate the mean of
- `trim` an argument that allows you to "trim" a fraction of the values
  from each end of the range of the variable prior to calculating the
  mean
- `na.rm` an argument that specifies whether missing values (labeled as
  "NA's") should be removed prior to calculating the mean or not — which
  can be specified with `na.rm = TRUE` or `na.rm = FALSE`

Importantly, the default `na.rm` argument setting in the `mean()`
function is `na.rm = FALSE`. Therefore, if you are calling the `mean()`
function with a variable that has missing values (i.e., "NA's"), you
would have to specify that by setting `mean(variable, na.rm = TRUE)`.
Otherwise, you will get an error when trying to use the function.

Even very experienced R users do not remember all of the various options
that are possible for the various arguments within functions. Calling
help on any function `?function.name()` to review the specific argument
possibilities available is a common activity for most users. Searching
for examples of how to use functions online, either on pages of official
documentation or otherwise, is also extremely common!

You could create your own functions in R fairly easily as well. As you
gain experience with the programming language, building your own custom
functions to enact common or repetitive tasks can save you a lot of time
and typing.

## *Installing and Loading Packages*

You can easily expand the functionality of R by installing and using
extension packages made by developers within the open-source R
community. For beginner R users, this can be somewhat confusing since
the combination of base functions with the possibilities of functions
from additional packages can offer many ways for conducting any given
procedure in R. Often, however, additional packages can make procedures
in R more user-friendly. Packages that are available for download and
installation are documented on the [R
Website](https://cran.r-project.org/).

Here is an example of downloading a package and loading it for use in R.
The
[`rstatix`](https://cran.r-project.org/web/packages/rstatix/index.html)
package was designed by Alboukadel Kassambara to make some of the most
common statistical analyses that social scientists use simple to execute
in R.

Using the base R `install.packages()` function will signal to R to find
the package online, download it, and install it for use in your R
system. For the **rstatix** packages, that would look like this:
`install.packages("rstatix")`.

You also need to "load" a package once in each R session to use the
functions within it — you can use the `library()` function to do that.
That makes all of the functions within a package available to you.
Loading the **rstatix** packages would look like this:
`library(rstatix)`.

As an example of the convenience and functionality that packages can
offer, consider that it is possible to calculate descriptive statistics
for a given variable in R using base functions such as: `mean()`,
`sd()`, `range()`, `min()`, `max()`, `median()`, and so on. But, there
are more convenient ways to get all of those summary statistics with
other packages. For example, both the **rstatix** package and the
**psych** package contain a single function for returning all of those
summary values.

Here is an example of calculating descriptive statistics summary with
base R functions vs. the `get_summary_stats()` function from the
**rstatix** package:

```{r base_rstatix, exercise = TRUE, exercise.eval = FALSE}
# Again, these first lines are just creating the dataframe for this example.
numeric_variable <- c(1.2, 3.7, 5, 2.2, 4.69, 6.123, 10.1, 4.3, 5.5, 6.09)
integer_variable <- c(4, 8, 15, 16, 23, 42, 12, 34, 78, 2)
character_variable <- c("who", "what", "when", "who", "what", "when", "who", 
                        "what",  "who", "what")
my_data <- data.frame(numeric_variable, integer_variable, character_variable)

# Base R functions for summary statistics
mean(my_data$numeric_variable)
sd(my_data$numeric_variable)
range(my_data$numeric_variable)
min(my_data$numeric_variable)
max(my_data$numeric_variable)
median(my_data$numeric_variable)

# First "load" the package
library(rstatix)

# Then use the "get_summary_stats()" function to get a variety of summary statistics.
get_summary_stats(my_data, numeric_variable)
```

### *Tidyverse*

The [**tidyverse**](https://www.tidyverse.org/) refers to a suite of 8
packages created by a number of developers led by Hadley Wickham that
are designed to provide more intuitive methods for conducting various
data science tasks in R. Some tidyverse packages we will use include
**ggplot2**: a graphing package, **dplyr**: a data manipulation package,
and **readr** as well as **haven**: packages for importing data.
Installing and loading **tidyverse** will set you up with all of the 8
packages at once.

The **rstatix** package demonstrated above is designed to work
seamlessly with other packages in the tidyverse as well.

## Importing and Viewing Various Types of Data Files

Conducting data analyses in R requires loading your data set as an
object in the R Global Environment. Unlike some programs such as SPSS or
Excel, there is not a direct data editor.

Four common types of data files you are likely to encounter include:

- CSV Files (XXX.csv): "Comma-separated-value" data that can be created
  in a variety of programs
- Excel Files (XXX.xlsx): Files create in Excel
- SPSS Files (XXX.sav): Data editor files created in SPSS (which can
  contain a good deal of meta-data, e.g., variable names, value labels,
  and so on)
- Stata Files (XXX.dta): Data files created in Stata (which can also
  contain meta-data)

The [**readr**](https://readr.tidyverse.org/) package includes a
function for importing csv files: `read_csv()`. The
[**readxl**](https://readxl.tidyverse.org/) package has a function for
important Excel files: `read_xls()`. The
[**haven**](https://haven.tidyverse.org/) package includes functions for
importing SPSS files: `read_sav()`, and Stata files: `read_dta()`.

Functions for reading in different types of data files have many
arguments for telling R how to interpret the file. We won't deal with
those here, suffice to say it's important to (a) know what's in the file
you are reading into R and (b) check that the file came in the way you
wanted it to.

The following code demonstrates using functions in the readr package to
load each of the data types described above. The data are files included
as a supplement to the Miksza & Elpus text, [*Design and Analysis for
Quantitative Research in Music
Education*](https://www.amazon.com/Design-Analysis-Quantitative-Research-Education/dp/0199391904)
(Chapter 4).

[Note: this code won't actually run, because importing external files
into this tutorial program was too hard for me to figure out how to do.]

```{r import_data, eval = FALSE}
# First we will load the packages with the necessary importing functions
library(readr)
library(haven)

# Assuming you are working within an R project and that your data files are in
# the same folder as your project files, importing the data can require very little code.
data_csv <- read_csv("Miksza & Elpus - Chorister data for descriptive analysis examples.csv")
data_SPSS <- read_sav("Miksza & Elpus - Chorister data for descriptive analysis examples.sav")
data_Stata <- read_dta("Miksza & Elpus - Chorister data for descriptive analysis examples.dta")

# Once imported you could use any number of functions to view the data, 
# e.g, "head()," "tail()," "str()," "glimpse()."
str(data_csv)
```

## Selecting, Filtering, and Altering Variables within Dataframes

In this section, we will use the `select()`, `filter()`, and `mutate()`
functions from the **dplyr** package.

Given that these functions are from a package within the tidyverse
ecosystem, they each can be used with what is referred to as a "pipe
operator": `%>%`. The pipe operator allows you to write code that can be
more intuitive to interpret because you can read it linearly, from left
to righ.

For these exercises, we will use a dataset called `DescriptiveData` with
four variables meant to represent characteristics of singers in a choir:
(a) `Age` — a numeric variable (i.e., ratio measure), (b)
`Years.experience.in.singing.ensemble` - an integer variable (i.e.,
ratio measure), (c) `Sex` - self-reported biological sex at birth, a
two-level factor variable (i.e., nominal measure), and (d)
`Preference.for.rehearsal.frequency` - a five-level ordered factor
variable (i.e., ordinal measure).

The first six rows of the data are shown below:

```{r show descriptive data, exercise = FALSE, exercise.eval = FALSE, echo = FALSE}
head(DescriptiveData) %>% kable(align = "c")
```

### *Selecting Variables*

First we start with the dataframe object then use the pipe `%>%` to
indicate that we want to pass the data to another function, then we call
the name of the variable we want to select within the `select()`
function, and then we can call a function that we would like to use to
act upon the selected variable - note that you don't need to actually
write the name of the variable in the final function.

```{r select, exercise = TRUE, exercise.eval = FALSE}
# This is an example of selecting the "Age" variable and then identifying the
# minimum age value (i.e., the youngest age).
DescriptiveData %>% select(Age) %>% min()
```

Try out the `select()` function and the `max()` function to identify the
largest `Years.experience.in.singing.ensemble` among the participants:

```{r select_user, exercise = TRUE, exercise.eval = FALSE}

```

### *Filtering Data*

Again, we start with the dataframe object then use the pipe `%>%` to
indicate that we want to pass the data to the following function, then
we call the name of the variable we want to use to `filter()` our data
along with the corresponding conditions. You can then continue with
another pipe to act upon the filtered data with additional functions, or
you could create a new dataset from the filtered data assigning the
entire line of code to a newly named object.

```{r filter1, exercise = TRUE, exercise.eval = FALSE}
# Here is an example of filtering the data so that only those who are 40 years or
# older remain, and then a call of the count() function to show how many of the 
# original 115 participants are left in the newly filtered data.
DescriptiveData %>% filter(Age >= 40) %>% count()

# Here is what it would look like to filter the data as above and then to assign
# the newly filtered data to a new dataframe object
DescriptiveData_Filtered <- DescriptiveData %>% filter(Age >= 40)

# You could use the "head()," "tail()," "glimpse()," or "str()" functions to view
# the new dataframe titled "DescriptiveData_Filtered." (That's not included here
# for the sake of saving space).
```

Here is an example of filtering the data so that only those indicating
`Sex` as Male are in the newly filtered data - note that quirks in this
code include that the "equals" operator requires two signs `==`, and
that when you work with levels of factor variables, they need to be
presented in quotes, and that the variable names and category labels are
case-sensitive.

```{r filter2, exercise = TRUE, exercise.eval = FALSE}
DescriptiveData %>% filter(Sex == "Male") %>% count()
```

Use the code space below to try out the `filter()` function and the
`count()` function to filter the data so that only those younger than
`Age` 40 are left in the data, then try to filter the data so that only
those indicating a biological sex as Female are in the newly filtered
data:

```{r filter_user, exercise = TRUE, exercise.eval = FALSE}

```

### *Creating New Variables*

Once again, we start with the dataframe object then use the pipe `%>%`
to indicate that we want to pass the data to the following function,
but, with `mutate()` we need to specify a name for our new variable and
then a formula for how create the new variable.

```{r mutate, exercise = TRUE, exercise.eval = FALSE}
# The code below creates a new variable named "Age_Months" that is each participants' 
# original Age value multiplied by 12, then the "head()" function is called so that
# you can see that the new variable is appended to the dataframe.
DescriptiveData %>% mutate(Age_Months = Age*12) %>% head()
```

Use the code space below to try out the `mutate()` function and the
`head()` function to create a variable that is each participants' `Age`
in days:

```{r mutate_user, exercise = TRUE, exercise.eval = FALSE}

```

## Basic Descriptive Analyses

### *Summary Statistics*

Creating frequency and crosstabs tables are perhaps two of the most
basic and fundamental tasks that analysts need to be able to do. The
base R functions `table()` and `proportions()` are good tools for these
tasks. The `table()` function will produce frequency counts in each
table cell whereas the `proportions()` function will produce cell
proportions.

```{r tables1, exercise = TRUE, exercise.eval = FALSE}
# Since "table()" is a base R function, we will not use the "pipe" as with 
# functions that work within the tidyverse. Instead, we'll use the dataframe and
# dollar sign approach to choose variables to analyze.

# Here is code to create a frequency table of the participants' preferred
# rehearsal frequency.
table(DescriptiveData$Preference.for.rehearsal.frequency)

# We can also assign the entire table to an object to make it easier to work
# with in other functions.
Rehearse_table <- table(DescriptiveData$Preference.for.rehearsal.frequency)

# The proportions function works with table objects.
proportions(Rehearse_table)

# If the number of decimal places is unwieldy, you can use the base R "round()"
# function to specify how many places you'd like to see.
round(proportions(Rehearse_table), 2)
```

Here is an example of a cross-tabulation using the same base R
functions.

```{r tables2, exercise = TRUE, exercise.eval = FALSE}
# Here is code to create a cross-tabulation table of the participants' preferred 
# rehearsal frequency according to biological sex.
table(DescriptiveData$Sex, DescriptiveData$Preference.for.rehearsal.frequency)

Bio_Rehearse_Table <- table(DescriptiveData$Sex, DescriptiveData$Preference.for.rehearsal.frequency)

round(proportions(Bio_Rehearse_Table), 2)
```

You could also request percentages *by row* by specifying an additional
argument within the `proportion()` function of `margin = 1` or *by
column* with `margin = 2`.

```{r tables3, exercise = TRUE, exercise.eval = FALSE}
Bio_Rehearse_Table <- table(DescriptiveData$Sex, DescriptiveData$Preference.for.rehearsal.frequency)

round(proportions(Bio_Rehearse_Table, margin = 1), 2)

round(proportions(Bio_Rehearse_Table, margin = 2), 2)
```

As mentioned on earlier pages, the **rstatix** package provides many
functions that make conducting common types of analyses convenient
within a tidyverse ecosystem. The `freq_table()` function can be used in
a similar way as the base R `table()` function. The `freq_table()`
function returns frequency counts and percentages (even the column is
labeled "prop").

```{r tables_rstatix1, exercise = TRUE, exercise.eval = FALSE}
# A line of code using a pipe operator can be used for this.

DescriptiveData %>% freq_table(Preference.for.rehearsal.frequency)
```

Here is an example of an **rstatix** cross-tabulation with the
`freq_table()` function.

```{r tables_rstatix2, exercise = TRUE, exercise.eval = FALSE}
# A line of code using a pipe operator can be used for this.

DescriptiveData %>% freq_table(Sex, Preference.for.rehearsal.frequency)
```

Examining descriptive statistics of quantitative variables is also a
fundamental step that all analysts will need to take. Such variables can
be summarized with the `get_summary_stats()` function from **rstatix**.

```{r summary_stat, exercise = TRUE, exercise.eval = FALSE}
# This code will provide summary statistics for the Age variable.

DescriptiveData %>% get_summary_stats(Age)
```

### *Summary Statistics by Sub-Group*

It is also often necessary to summarize data according to sub-groups.
The `group_by()` function from the **dplyr** package can be used for
that type of task.

```{r summary_stat_group, exercise = TRUE, exercise.eval = FALSE}
# This code will provide summary statistics for the Age variable broken out by
# participants' reported biological sex

DescriptiveData %>% group_by(Sex) %>% get_summary_stats(Age, type = "common")
```

### *Correlation*

Examining bivariate correlations is also a common descriptive analysis
task. As with all things in R, there are many, many ways to calculate
correlation coefficients and a variety of packages with correlation
functions. The **rstatix** package has several useful functions for
calculating correlations. An additional element of flexibility in all of
the rstatix correlation functions is the ability to specify among
several common types: Pearson, Spearman, or Kendall.

```{r correlate1, exercise = TRUE, exercise.eval = FALSE}
# The "cor_test()" function is quite flexible and intuitive. This code returns
# a Pearson correlation between Age and Years experience in a singing ensemble.

DescriptiveData %>% cor_test(Age, Years.experience.in.singing.ensemble)
```

There is also a useful function in the **rstatix** package for creating
correlation matrices `cor_mat()`. Since there are only two
interval/ratio level variables in this example dataset, it would be
inappropriate to demonstrate a Pearson correlation matrix of more than
two variables.

It could also be helpful to be able to examine correlations according to
sub-groups. This code first filters the `Sex` variable so only those who
reported their biological sex are considered in the analyses, then it
will result in the Pearson correlation between `Age` and
`Years.experience.in.singing.ensemble` broken out by participants'
reported biological `Sex`.

```{r correlate2, exercise = TRUE, exercise.eval = FALSE}
DescriptiveData %>% filter(Sex == "Male" | Sex == "Female") %>%
  group_by(Sex) %>% cor_test(Age, Years.experience.in.singing.ensemble)
```

## Basic Data Visualization with ggplot2

R's graphic capabilities are amazing because the options for visualizing
data are vast. In this section, we will again use the `DescriptiveData`
example dataframe to demonstrate how to make some of the most common
types of graphs. We will continue working within the **tidyverse**
ecosystem and use the extremely flexible package **ggplot2**.

### *ggplot2*

The "gg" in **ggplot2** stands for the "grammar of graphics," which is a
reference to a book by Leland Wilkinson about coding programs for data
visualization. The primary developer of **tidyverse** packages and
functions, Hadley Wickham, also developed **ggplot2**. The most basic
elements or "layers" of a plot rendered with **ggplot2** include: (a) a
dataframe, (b) the designation of "aesthetics" (e.g., the mapping of
variables to axes and aesthetic features), and (c) a "geom" (e.g., a
shape for the plot such as histogram, boxplot, barplot, scatterplot, and
so on).

From there, there are many options for building additional layers to the
plot and a tremendous amount of flexibility in determining the specific
look of the features of the plot. When it comes to making beautiful
plots, there are too many good books and tutorials online that you can
learn from. For our purposes in this section, we will just make basic
(albeit ugly) plots. Several options for customizing plots will be
demonstrated in the next section.

The most basic code for a plot would look something like this:

- First call the function `ggplot()` and specify the data source:
  - `ggplot(data = your_data)`
- Next, specify which variables from the dataframe are assigned to which
  axes using the `aes()` function inside of the `ggplot()` function:
  - `ggplot(data = your_data, aes(x = variable_1))`
  - (There are many more aesthetic possibilities beyond designating
    axes)
- Next, choose a "geom" to indicate what type of plot you are aiming to
  make by adding one of the possible options with a plus sign (`+`)
  after the `ggplot()` function:
  - `ggplot(data = your_data, aes(x = variable_1)) + geom_histogram()`

### *Histogram*

Here is an example of code for creating a histogram that follows the
generic form described above for the `Age` variable from the example
`DescriptiveData` dataset.

```{r histogram, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# Note that the "data =" part of the code was removed since the ggplot() function
# will assume the first argument is the name of the dataframe.

ggplot(DescriptiveData, aes(x = Age)) + geom_histogram()
```

Use the code space below to try and code a histogram for the
`Years.experience.in.singing.ensemble` variable:

```{r histogram_user, exercise = TRUE, exercise.eval = FALSE, message = FALSE}

```

### *Boxplot*

Creating a boxplot instead is simply a matter of choosing a different
geom. Here is an example of code for creating a boxplot of the `Age`
variable. Note that the values on the y axis are not informative and
refer to the width of the box. You'll see how to remove them in the next
section of this tutorial.

```{r boxplot1, exercise = TRUE, exercise.eval = FALSE}
ggplot(DescriptiveData, aes(x = Age)) + geom_boxplot()
```

You could make it a more typical looking vertically-oriented boxplot by
assigning the `Age` variable to y instead of x.

```{r boxplot2, exercise = TRUE, exercise.eval = FALSE}
ggplot(DescriptiveData, aes(y = Age)) + geom_boxplot()
```

Use the code space below to try and code a boxplot for the
`Years.experience.in.singing.ensemble` variable:

```{r boxplot_user, exercise = TRUE, exercise.eval = FALSE}

```

### *Barplot*

Creating barplots for nominal (i.e., factor in R) or ordinal variables
(i.e., ordered factors in R) can be done with very similar code.

```{r barchart, exercise = TRUE, exercise.eval = FALSE}
ggplot(DescriptiveData, aes(x = Preference.for.rehearsal.frequency)) + geom_bar()
```

Creating "stacked" or "dodged" barplots for two nominal (i.e., factor in
R) or ordinal variables (i.e., ordered factors in R) would require
assigning a second variable to `fill = variable_2` in the `aes()`
function and then specifying an additional argument of
`position = "stack"` or `position = "dodge"` within the `geom_bar()`
function.

Note that the axis labels are very crowded in this plot. We will look at
making plot customizations to address issues like that in the next
section.

```{r barchart_dodge, exercise = TRUE, exercise.eval = FALSE, message =FALSE}
ggplot(DescriptiveData, aes(x = Preference.for.rehearsal.frequency,
                            fill = Sex)) + geom_bar(position = "dodge")

# Note that this plot is particularly odd since there are "empty cells" for 
# some combinations of these two variables.
```

### *Scatterplot*

Creating scatterplots to depict the relationship between two
quantitative variables will feel similarly intuitive once you get the
hang of the grammar of graphics approach. Here is a scatterplot of the
`Age` and `Years.experience.in.singing.ensemble` variables.

```{r scatterplot1, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(DescriptiveData, aes(x = Age,
                            y = Years.experience.in.singing.ensemble)) + geom_point()
```

Adding a layer consisting of a line of best fit and the associated
standard error can be done by adding another layer with `geom_smooth()`
along with an argument specifying a linear model: \`method = "lm".
(Clearly this is a very weak/non-existent correlation!)

```{r scatterplot2, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(DescriptiveData, aes(x = Age,
                            y = Years.experience.in.singing.ensemble)) + 
  geom_point() +
  geom_smooth(method = "lm")
```

## A Bit Less Basic Data Visualization with ggplot

In this section, a few of the additional options for customizing plots
in ggplot2 will be demonstrated. We will use a different dataset that is
built into this package for these demonstrations since it is larger and
can allow for few more options. It is titled `InferentialTestData`.

This dataset contains nine variables representing characteristics and
measurements of school children: (a) `mathtestk` - kindergarten
students' standardized math test scores, a numeric variable (i.e.,
interval measure), (b) `mathtest1` - those same students' 1st grade
standardized math test scores, a numeric variable (i.e., interval
measure), (c) `urbanicity` - the urbanicity of the students' school
(i.e., city, suburb, town, rural), a factor variable (i.e., nominal
measure), (d) `music` - whether they participated in school music class
(i.e., No Music, Yes Music), a factor variables (i.e., a nominal
variable), (e) `ses` - a quantitative indiactor of students'
socioeconomic status, and (f) a set of "dummy" variables that indicate
"Yes" or "No" for each urbanicity category - factor variables. This
particular dataset is used to illustrate analyses in "Chapter 10:
Regression" of the Miksza & Elpus text, [*Design and Analysis for
Quantitative Research in Music
Education*](https://www.amazon.com/Design-Analysis-Quantitative-Research-Education/dp/0199391904).

The first six rows of the data are shown below:

```{r show inferentialTest data, exercise = FALSE, exercise.eval = FALSE, echo = FALSE}
head(InferentialTestData) %>% kable(align = "c")
```

### *Additional Common Plot Customizations*

First, we will consider methods for enacting several very common
desirable alterations to default plots generated by **ggplot2**. Then,
we will create some plots incorporating these methods. Please be aware
that these tips barely scratch the surface of the customizations
available in the **ggplot2** package. Fortunately, there are many, many
terrific online resources for making virtually any sort of alteration
you'd like.

#### *Additional Aesthetic Mappings*

All primary elements of a plot can be customized according to how you
would like them to appear. For example, within the `aes()` function it
is possible to specify arguments that control outline and fill colors,
point size, point shape, line thickness, linetype,degree of
transparency, geom dimensions, and so on. Various aesthetic mappings
will be demonstrated in the sample code presented below.

#### *Axis Labels*

By default, `ggplot()` will use variable names as axis label and will
not include any additional title or sub-title information. Adding custom
labels and titles can be done by adding an additional function to the
plot code of `+` `labs()` and specifying labels in quotes with arguments
in the `labs()` function such as: `x =`, `y =`, `title =`, `subtitle =`,
or `caption =`.

#### *Adjusting Axis Ranges and Ticks*

Sometimes when plotting quantitative variables, the default plot will
produce axes and indicators or "ticks" of values that make it
challenging to interpret the range of values in a plot. You can
customize x and y axes in `ggplot()` by adding additional functions such
`+` `scale_x_continuous()` and `+` `scale_y_continuous()`. You can
specify the range of values to be included by inserting the arguments
`limits =` into the previous functions, followed by a vector of two
values representing the desired minimum and maximum, such as, `+`
`scale_x_continuous(limits = c(1, 100))` - that would create an x axis
bound by 1 and 100.

If, say, you'd like to customize the axis ticks that are shown between
the axis minimum and maximum values, then you could also insert the
argument `breaks =` into the `scale_x_continuous()` function, followed
by a vector of values specifying each tick point. For example, the
following code would produce an x axis with a range of 1 to 100 and
ticks at every multiple of 10:

- `scale_x_continuous(limits = c(1, 100), breaks = c(10, 20, 30, 40, 50, 60, 70, 80, 90, 100))`.

There are also similar methods available for customizing axes
representing all sorts of variable types, not just quantitative
variables.

#### *Theme Elements*

The general "look" of a plot can be controlled with a number of
arguments that can be specified within a `+` `theme()` layer. Take a
look at the online documentation of the various elements that can be
adjusted within a theme layer: [Theme Element
Options](https://ggplot2.tidyverse.org/reference/theme.html). The
options are certainly numerous enough that they can be overwhelming when
getting started.

However, **ggplot2** also has several complete themes that you could use
which convert many elements of a plot to a particular style. You can see
some of the options for complete themes at the *tidyverse* website as
well: [Complete Theme
Options](https://ggplot2.tidyverse.org/reference/ggtheme.html). The name
of the the complete theme is typically appended to the `ggplot()`
function. For example, to use what is referred to as a minimalist style
in `ggplot()` you could add a layer that looks something like: `+`
`theme_minimal()`.

### *Histogram*

Here is code for a histogram depicting the distribution of the students'
1st grade standardized math test score variable.

```{r histogramDetail1, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(InferentialTestData, aes(x = mathtest1)) + geom_histogram()
```

Here is code for the same histogram, but this time incorporating the
features described above. Note that you can continue adding layers to
the plot code on a new line if the previous line ends with a `+` sign.

```{r histogram_userDetail2, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(InferentialTestData, aes(x = mathtest1)) + 
  geom_histogram(color = "white", fill = "blue") +
  labs(title = "Distribution of 1st Grade Students' Standardized Math Test Scores",
       x = "1st Grade Math Test Score", y = "Frequency") +
  scale_y_continuous(limits = c(0, 100), breaks = c(20, 40, 60, 80, 100)) +
  theme_minimal()
```

Use this code space to customize your own histogram of `mathtest1`
scores.

```{r histogramDetail_user, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
```

### *Boxplot*

Here is code for a vertical boxplot depicting the distribution of the
students' 1st grade standardized math test score variable.

```{r boxplot1Detail1, exercise = TRUE, exercise.eval = FALSE}
ggplot(InferentialTestData, aes(y = mathtest1)) + geom_boxplot()
```

Here is code for the same boxplot, but this time incorporating the
features described above with customizations. We will add the
traditional "whiskers" to this version with the additional argument of
`staplewidth =` within the `geom_boxplot()` function and eliminate the
nuisance information on the x axis by specifying `breaks = NULL` in the
`scale_x_continuous()` function.

```{r boxplot1Detail2, exercise = TRUE, exercise.eval = FALSE}
ggplot(InferentialTestData, aes(y = mathtest1)) + 
  geom_boxplot(fill = "lightblue", staplewidth = .25) +
  labs(title = "Distribution of 1st Grade Students' Standardized Math Test Scores",
       y = "Math Test Score") +
  scale_x_continuous(breaks = NULL) +
  scale_y_continuous(limits = c(0, 100), breaks = c(0, 15, 30, 45, 60, 75, 90)) +
  theme_classic()
```

Use this code space to customize your own boxplot of `mathtest1` scores.

```{r boxplot_userDetail, exercise = TRUE, exercise.eval = FALSE}
```

### *Barplot*

Here is code for a barplot depicting `urbanicity` variable. Note that
the labels for each bar are taken directly from the factor variable
levels. You can change those by relabeling the levels of factor variable
outside of the `ggplot()` function using the `labels =` argument in the
`factor()` function.

```{r barchartDetail, exercise = TRUE, exercise.eval = FALSE}
ggplot(InferentialTestData, aes(x = urbanicity)) + geom_bar()
```

Here is code for a dodged barplot depicting the frequency of students in
each urbanicity category according to whether they are in school music
or not. Note that the title and labels for the legend that is produced
are drawn from the name of the music factor variable and names of its
levels. These can be adjusted by renaming the variable and relabeling
the factor levels.

```{r barchart_crosstabDetail1, exercise = TRUE, exercise.eval = FALSE}
ggplot(InferentialTestData, aes(x = urbanicity, fill = music)) + 
  geom_bar(position = "dodge")
```

Here is code for the same barplot with additional customizations. Note
that with two variables, you can also specify the color scheme with more
specific code. In this example, the `scale_fill_manual()` function is
used to adjust the look of the `fill =` aesthetic. You will find similar
functions for other aesthetic mappings (`scale_color_manual()`,
`scale_linetype_manual()`, etc.).

```{r barchart_crosstabDetail2, exercise = TRUE, exercise.eval = FALSE}
ggplot(InferentialTestData, aes(x = urbanicity, fill = music)) + 
  geom_bar(position = position_dodge(width = 1), width = .8) +
  scale_fill_manual(values = c("red", "blue")) +
  labs(x = "Urbanicity", y = "Frequency Count",
       title = "Cross-tabulation of Urbanicity and School Music Participation",
       subtitle = "This is an example of a sub-title.") +
  theme_minimal()
```

Use this code space to customize your own dodged barplot.

```{r barchart_crosstabDetail_user, exercise = TRUE, exercise.eval = FALSE}
```

### *Scatterplot*

Here is code for a scatterplot depicting the relationship between the
students' kindergarten and 1st grade standardized math test scores.

```{r scatterplot1Detail1, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(InferentialTestData, aes(x = mathtestk, y = mathtest1)) + 
  geom_point()
```

Here is code for the same scatterplot with customizations. Note that the
`alpha =` argument is used in the `geom_point()` function to add a
degree of transparency to the points so that the overplotting (e.g.,
lots of points in the same general area) is a bit easier to see. Also,
since this is the final example in this section, it demonstrates
including a variety of arguments that can be used for specific theme
specifications rather than a complete theme.

```{r scatterplot1Detail2, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
ggplot(InferentialTestData, aes(x = mathtestk, y = mathtest1)) + 
  geom_point(color = "blue", alpha = .35) +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(x = "Kindergarten Math Test", y = "1st Grade Math Test",
       title = "Scatterplot of Participants' Kindergarten and 1st Grade Test Scores") +
  scale_x_continuous(limits = c(5, 85), 
                     breaks = c(10, 20, 30, 40 , 50, 60, 70, 80)) +
  scale_y_continuous(limits = c(5, 85),
                     breaks = c(10, 20, 30, 40 , 50, 60, 70, 80)) +
  theme(title = element_text(size = 15),
        axis.line = element_blank(),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 13),
        panel.grid.minor = element_line(color = "lightgrey"),
        panel.grid.major = element_line(color = "grey"),
        panel.background = element_rect(fill = NA),
        axis.ticks = element_blank())
```

Use this code space to customize your own scatterplot.

```{r scatterplotDetail_user, exercise = TRUE, exercise.eval = FALSE}

```

## Basic Inferential Analyses and Formula Notation

We will continue to use the `InferentialTestData` dataset described in
the previous section for the examples in this section. To keep things
relatively simple, we will also restrict our analyses to the use of the
base R linear model function, i.e., `lm()`, and functions available in
the **rstatix** package described earlier.

In addition, all of the functions that will be demonstrated in the
remaining sections can take a formula as the argument for specifying the
test you would like to conduct. As such, we will begin with a quick
introduction to formula notation.

### *Using Formula Notation*

For the analyses demonstrated in the next section, the formula notation
will involve specifying a dependent variable (i.e., an outcome, `y`) and
independent variables (i.e., predictors, `x`). The tilde, `~`, symbol is
used to separate the dependent from the independent variables.
Therefore, a basic analysis of the influence of one independent variable
on one dependent variable would look like this:

|   `y ~ x`

\
For analyses with more than one independent variable, a plus sign, `+`,
can be used to extend the formula. If interaction effects are desired
between independent variables, then a colon, `:`, could be used between
them. Therefore, an analysis involving two independent variables (or an
independent variable and a covariate, etc.) could look like this:

|   `y ~ x1 + x2`

The interaction effect could be added like this:

|   `y ~ x1 + x2 + x1:x2`

### *A Quick Note About Options Available in R*

It's worth taking a moment to mention that in addition to base R
functions, there are many packages that have functions for doing the
relatively simple analyses presented in the next sections.

Also recall that the **rstatix** package is designed to work well with
the **tidyverse** suite of packages. As such, we will make frequent use
of the pipe operator, `%>%` to pass the `InferentialTestData` dataframe
to various tests functions.

## Linear Regression Models

Many of the statistical tests arts education researchers wish to use can
be executed with some generalized form of the **general linear model**.
For example, correlation, t-tests, ANOVAs, and regression analyses can
all be conducted with the same function, `lm()`.

The `lm()` function is quite simple to use. It involves specifying an
argument that is a formula with variable names to indicate dependent and
independent variables (e.g., `outcome_var ~ predictor_var`) and then
specifying an argument indicating the dataframe in which those variables
can be found (e.g., `data = your_dataframe`).

### *Simple Linear Regression*

#### *Interval/Ratio Predictor Variables*

Here is an example of specifying a linear model that analyzes the
relationship between participants' kindergarten and 1st grade
standardized math test scores. It is helpful to assign the results of an
`lm()` analysis to its own object. That way, you can call the base R
`summary()` function on the `lm()` object, which will yield an
informative results output.

With two numeric variables, this is essentially the same as conducting a
Pearson correlation. In fact, calling for the "standardized" regression
coefficient for this analysis will result in the Pearson coefficient.
There is a function for standardized regression coefficients in the
**QuantPsyc** package called `lm.beta()` that can be used for this.

```{r regression_cor, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First, specify the model and assign it to an object with a name you make up.
kind_1st_mod <- lm(mathtest1 ~ mathtestk, data = InferentialTestData)

# Next, call "summary()" to get the results.
summary(kind_1st_mod)
```

Evaluating the assumptions of a linear model is also a straightforward
task in R. Calling the generic base R `plot()` function on a `lm()`
object will result in a series of plots that can be used to assess the
heterescedasticity and normality of residuals.

```{r regression_cor_assump, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First, specify the model and assign it to an object with a name you make up.
kind_1st_mod <- lm(mathtest1 ~ mathtestk, data = InferentialTestData)

# Use the "plot()" function to evaluate model assumptions
plot(kind_1st_mod)
```

For illustration purposes, here is code calling the `lm.beta()` function
from the **QuantPsyc** package on the model to get the standardized
coefficient. This is followed by code for the Pearson correlation for
confirmation of the equivalence.

```{r regression_cor_stand, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# Getting the standardized coefficient
lm.beta(lm(mathtest1 ~ mathtestk, data = InferentialTestData))

# Pearson correlation
InferentialTestData %>% cor_test(mathtest1, mathtestk)
```

#### *Nominal/Ordinal Predictor Variables*

Assuming the variables in a dataframe are appropriately classified
according to their measurement scale (e.g., numeric, factor, ordered
factor), then R will interpret them appropriately in the within `lm()`
function. Therefore, using nominal or ordinal predictor variables will
result in R automatically creating k-1 dummies with the 1st level of the
factor as the reference category for such variables. There are fairly
simple ways to change the reference category and the general approach
for dealing with factor variables within `lm()`, but we won't cover
those today.

Here is an example of specifying a linear model that analyzes whether
participants' kindergarten standardized math test scores varies as a
function of the urbanicity of their school. Note that the reference
category for the k-1 `urbanicity` dummy variables is the "city"
category.

With an interval/ratio-level outcome variable and a more than 2-category
nominal predictor variable, this is essentially the same as conducting
an ANOVA.

```{r regression_nom, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First, specify the model and assign it to an object with a name you make up.
mathk_urb_mod <- lm(mathtestk ~ urbanicity, data = InferentialTestData)

# Next, call "summary()" to get the results.
summary(mathk_urb_mod)
```

You can actually call the base R `anova()` function on an `lm()` object
to get the traditional ANOVA table as output rather than regression
results.

```{r regression_nom_anova, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First, specify the model and assign it to an object with a name you make up.
mathk_urb_mod <- lm(mathtestk ~ urbanicity, data = InferentialTestData)

# Next, call "anova()" to get an ANOVA table.
anova(mathk_urb_mod)
```

### *Multiple Linear Regression*

Adding multiple predictor variables/covariates or higher order terms
(e.g., a squared term for quadratic effects) is simply a matter of
continuing the formula argument with plus signs `+` and additional
variable names. As mentioned above including interaction terms can be
done with the `:` symbol between variable names, as in
`variable1:variable2`.

Here is code for a model that regresses both, whether participants
participated in music and their socioeconomic status on their 1st grade
standardized math test scores.

```{r regression_mult, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First, specify the model and assign it to an object with a name you make up.
math1_mult_mod <- lm(mathtest1 ~ music + ses, data = InferentialTestData)

# Next, call "summary()" to get the results.
summary(math1_mult_mod)
```

#### *Model Comparison*

Model comparisons can be conducted by assigning the models you would
like to compare to separate objects (e.g., `model_1`, `model_2`), and
then calling the `anova()` function with the two model names as the
arguments (e.g., `anova(model_1, model_2)`).

```{r regression_mult_compare, exercise = TRUE, exercise.eval = FALSE, message = FALSE}
# First specify the single predictor model.
math1_mus_mod <- lm(mathtest1 ~ music, data = InferentialTestData)

# Next specify the multiple predictor model.
math1_mult_mod <- lm(mathtest1 ~ music + ses, data = InferentialTestData)

# Finally, call "anova()" function to compare the two models.
anova(math1_mus_mod, math1_mult_mod)
```

## Traditional Tests for Group Comparisons

### *Comparing Two Independent Samples*

The following code uses the `t_test()` function to compare 1st grade
standardized math test scores according to whether the participants were
in music or not.

```{r independent_t, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% t_test(mathtest1 ~ music)
```

If the variances of our groups were not equal, we could adjust the
default setting of the `var.equal =` argument within the `t_test()`
function to be `var.equal = FALSE`. By doing that, the more robust Welch
version of the t-test would be run.

If we needed to compare "dependent samples" instead, then we could
adjust the `paired =` argument from its default setting of `FALSE` to
`paired = TRUE`.

Use the code space below to test whether participants' socioeconomic
status, `ses` (i.e., a dependent variable in this case, `ses`) varies as
a function of whether participants were in music or not, `music` (i.e.,
an independent variable, `music`).

```{r independent_tuser, exercise = TRUE, exercise.eval = FALSE}
```

### *Effect Size for Comparing Two Independent Samples*

Calculating the effect size for the difference between two independent
means (i.e., Cohen's d) involves similar code with the `cohens_d()`
function:

```{r cohen_d, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% cohens_d(mathtest1 ~ music)
```

The same options to arguments described above can be made in the
`cohens_d()` function if the analyses involved independent samples with
unequal variance (`var.equal = FALSE`) or paired samples
(`paired = TRUE`). Similarly, an argument specifying
`hedges.correction = TRUE` can be added if Hedge's G is preferred.

Use the code space below to determine the effect size of the difference
in participants' socioeconomic status, `ses`, as a function of whether
participants were in music or not, `music`.

```{r cohen_duser, exercise = TRUE, exercise.eval = FALSE}
```

### *Comparing More than Two Independent Samples*

Conducting a oneway ANOVA to compare more than two independent samples
is also quite straightforward. Note that the effect size for this
analysis, in this case generalized eta squared, is returned from this
function by default.

```{r anova, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% anova_test(mathtestk ~ urbanicity)

```

Use the code space below to run a similar test, but with 1st grade
standardized math test scores, `mathtest1` as the dependent variable.

```{r anova_user, exercise = TRUE, exercise.eval = FALSE}

```

Probing pairwise comparisons as a follow-up to ANOVA analyses can be
done in several ways. Here we use the `tukey_hsd()` function. Note that
this function also takes a formula and that the results returned are
adjusted for multiple comparisons using the Tukey method.

```{r anovapost, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% tukey_hsd(mathtestk ~ urbanicity)

```

Alternatively the `pairwise_t_test()` function with an additional
argument of `p.adjust.method =` set to one of the available options
(`"bonferroni"`, `"hochberg"`, and so on) could be used.

### *Factorial Designs*

ANOVA models involving more than one independent variable (and
interactions) can be estimated with the same `anova_test()` function
used above or with the `factorial_design()` function.

Here is an example of a test of whether 1st grade standardized math test
scores vary as a function of (a) music participation - `music`, (b)
urbanicity - `urbanicity`, or (c) whether the effect of music
participation depends upon students' urbanicity, that is, the
interaction of `music` and `urbanicity`.

(Also note that the argument `type = 3` was added to this code to
indicate the analysis should use type 3 sums of squares. Type 3 sums of
squares is a default in some commercial statistical packages such as
SPSS.)

```{r anova2way, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% anova_test(mathtest1 ~ music + urbanicity + music:urbanicity, type = 3)

```

Further pairwise comparisons of significant effects can be conducted in
a variety of ways. As one example, it might be desirable to see if
urbanicity is a significant effect for those who do and those who do not
participate in music. That can be done by first grouping according to
music, and then conducting tests of estimated marginal means with **p**
values adjusted for multiple comparisons.

```{r emmeans, exercise = TRUE, exercise.eval = FALSE}
InferentialTestData %>% group_by(music) %>%
  emmeans_test(mathtest1 ~ urbanicity, p.adjust.method = "bonferroni")

```

### *Specifying Within-Subjects Factors*

All of the ANOVA examples so far have involved between-subjects factors,
exclusively. Specifying models with within-subjects factors is a bit
more complicated in R. Exploring the options for such models is a bit
beyond the scope of this tutorial.

### *Assumptions and Non-Parametric Analogs*

Typical tests of assumptions for the tests mentioned above are also
available in the **rstatix** package and can be called with similar
code. Here are a few among the options available:

- levene_test()
- shapiro_test()
- identify_outliers()

The non-parametric analogs of the t-test and ANOVA for nominal or
ordinal outcome data are similarly accessible in the **rstatix**
package. Use the `?function_name()` help call on the tests below to see
what options are avaialble:

- `chisq_test()`
- `mcnemar_test()`
- `wilcox_test()`
- `kruskal_tes()`
- `friedman_test()`
